\section{Regressors}
\label{sec:modelling:regressors}
As discussed extensively already, the selection of controls is crucial to avoid unmeasured confounders. However,
a second important consideration is the action of mediators. Mediators are variables that explain the effect
of $x$ on $y$, in the sense that $x$ causes some mediator $m$ which causes $y$. An example of a mediator in the present
study might be physical health. If physical health is included as a control in the regression, the interpretation of $\beta$
becomes ``the effect of sports on MHI5-score, \textit{given} the physical health status of an individual.'' However,
improvement of physical health may well be a strong mechanism through which sports affects mental health, which
should very much be included in $\beta$ for the purpose of answering the research question. The subtleties of mediation
are often not discussed in the literature \cite{imbens2024causal}, e.g. \citeA{chekroud2018association} simply include physical health as a control.
Such mediation might be measured directly while still effectively controlling for confoundedness due to the mediator as
in the following, simplified example:
\begin{align}
\begin{split}
    \label{eq:modelling:mediation_example}
    y_{it} &= \alpha_y + \beta x_{it} + \eta m_{it} + \epsilon_y; \\
    m_{it} &= \alpha_m + \zeta x_{it} + \epsilon_m.
\end{split}
\end{align}
Then, the total effect of $x$ on $y$ is given as $\beta + \zeta \eta$, the sum of the direct and the indirect effects.
Conveniently, these two regressions may be estimated simultaneously in SEM.

Thus, when determining the controls used, attention must be paid to which variables may be mediators and those should be
modelled as in \cref{eq:modelling:mediation_example}. There is a significant price to be paid in terms of model parsimony
though, as a mediator is necessarily an endogenous variable, adding many additional parameters to the model.
If a mediator is simply excluded from the regression altogether, this model complexity is avoided, while the indirect
effect (e.g. $\zeta \eta$) is absorbed into the main parameter ($\beta$). However, in doing so, the mediator, in so far
as it is not fully explained by $x$, also functions as a potential unmeasured confounder.
Because it is not clear a priori which approach is preferred, the results for both will be given in the present study.
In this way, it can also be directly examined to which degree the mediators are confounders.

As covariates, we include a basic set of sociodemographic variables available in the Background Variables of the LISS panel,
namely age, ethnicity, gender, marital status, education level, employment status and net household income.
Of these, the last four might be hypothesised as mediators based on the literature
(\citeA{spiker2014mental,hjorth2016mental,frijters2014effect,thomson2022income} respectively).
However, all of these variables were found to vary so little in time that including them as time-variant covariates caused
numerical instability, so they could only be included as time-invariants. This precludes modelling them as mediators,
but the fact that they vary so little indicates that their effect as mediators could only be weak ($\zeta$ must be small),
so the error in estimating the total effect due to not modelling them as mediators could only be marginal.
Beyond this set, based on the literature and availability in the data, we include (self-reported) physical health
and disease status as additional mediators \cite{westcott2012resistance}.
While the Health study also queries more nuanced variables like hospital admittance and medication use, these are
highly correlated with health and disease status (e.g. $\rho \approx 0.55$ for medication use and disease status),
and are thus excluded for the sake of parsimony.

Because only the long-term effect of sports on mental health is considered (recall \cref{sec:methods:reverse_causality}),
it would be natural to the cumulative years that someone has exercised as the variable of interest, rather than the
yearly exercise status. However, because it is not known how many years one has exercised at the time of joining this panel,
this variable can only be roughly approximated through a running sum. Additionally, information missing in one year would
complicate using the variable in all future years.
As such, sports is used directly, while examining different orders of the distributed lags to model effects beyond
one year ahead.

This study chooses to not include individual-specific effects $\alpha_i$ in the model. This is because while they help justify
the ignorability assumption through eliminating unmeasured confounders, they also potentially obscure the effects of time-invariant mediators,
which equally invalidate the findings. The approach is thus instead to model individual-specific variability through the
time-invariant controls $z_i$. Additionally, inclusion of fixed effects entails estimating $N$ additional parameters
(the incidental parameter problem). Since it is likely that $N > \frac{p(p + 1)}{2}$, the model would not be identified.
That being said, nothing prevents modelling random effects. In fact, this is commonly done in
SEM, see for instance \citeA{heck2001multilevel}. However, the author leaves it as an avenue for further research.

As a minor final comment, some dummy levels of controls were excluded, because they occurred so infrequently in the data
that including them as regressors led to numerical instability.
These were a gender of ``other'' and, perhaps surprisingly, a household income between 15\,000 and 50\,000.

\section{Lag Selection}
\label{sec:modelling:lags}

\section{Fixing Parameters across Time}
\label{sec:modelling:parameter_fixing}
